---
title: "Untitled"
author: "Andrew Beveridge"
date: "9/23/2021"
output: html_document
---



```{r, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(mosaic))
```



The exam will have:


1. Power Function Zipf law (fit power function, thinking about models and conclusions)

2. Exp/Log Beer froth decay (exponential function from 2 points)? ACTUALLY GIVE A TABLE OF DATA INSTEAD

3. Periodic Measels in the 1950s periodic


4. Multivariable Drug Surge

5. Dimensional Analysis Object in fluid dimensional analysis

6. Fermi Problem

7. Linear Function: locational rent = gross income - production costs - transport costs.

8. Rate of change: MSP unemployment


linear functions
rate of change
multivariable functions

units
dimension
fermi

Lava java sketch the contour (pick from choices), estimate rate of change
or Kristin's Cobb-Douglas function about diminishing returns

Coffee consumed by Macalester students in a semester?

```{r}
2000 * 1.5  * 4*30 * 16 / 128
```


Pouring a beer into a mug creates a froth of foam. The height h(t) of the froth experiences continuous exponential decay where $i$ is measured in seconds. Find the rate $r$ of exponential decay for a beer that starts with a froth of height 12.35 mm that decays to height 5.02 mm after 60 seconds. Your numeric answer must use 3 decimal places.


```{r}
log(5.02/12.35)/60
exp(-0.0150)
0.985^60 * 12.35
```



```{r}
plotFun(35.74+0.6215*T-35.75*V^(0.16)+0.4275*T*V^(0.16)~T & V, T.lim=range(-30,30),
V.lim=range(5,40),zlab="V",xlab="T",ylab="W")
```
```{r}
tdata = seq(0.05,.5, .05)
distdata = c(0.3, 1.25, 2.4, 4.6, 7.1, 10, 13.8, 18.1, 22.6, 28.0)
plotPoints(distdata~tdata)
logtdata = log(tdata)
logdistdata = log(distdata)
fitModel(logdistdata~A*logtdata+B)
plotFun(exp(4.68) * x^1.97~x, add=TRUE, col='red')
```
```{r}
plotFun(exp(4.68) * x^1.97~x,  col='red')
```

```{r}
x = seq(1,14)
pop=c(8.80, 3.90, 2.75, 2.31, 1.61, 1.60, 1.43, 1.39, 1.30, 1.01, 0.96, 0.95, 0.92, 0.91)
plotPoints(log(pop) ~ log(x))
fitModel(log(pop)~A*log(x)+B)
plotPoints(pop~x)
plotFun(exp(2.007)*x^(-0.84)~x, add=TRUE, col='red')
```

###### Zipf Law

It has been long observed that the populations of the largest cities seem to follow a power law
$pop(r) = C r^p$ where $r$ is the rank of the city's population (1st, 2nd, 3rd, 4th largest, and so on). Find the values of C and p for US Metro Areas

```{r}
x = seq(1,14)
pop=c(18.82, 12.46, 8.88, 6.49, 6.40, 6.17, 5.91, 5.73, 5.38, 4.58, 4.32, 3.53, 3.46, 3.31)
plotPoints(log(pop) ~ log(x))
fitModel(log(pop)~A*log(x)+B)
exp(2.9139173)
exp(2.91)
plotPoints(pop~x)
plotFun(exp(2.91)*x^(-0.62)~x, add=TRUE, col='red')
```




Find the values of C and p for World Metro Areas

```{r}
x = seq(1,14)
pop=c(37.34, 31.18, 27.80, 22.24, 21.92, 21.74, 21.32, 20.90, 20.67, 19.11, 18.82, 16.50, 16.38, 15.42)
plotPoints(log(pop) ~ log(x))
fitModel(log(pop)~A*log(x)+B)
exp(3.6339626)
plotPoints(pop~x)
plotFun(exp(3.62)*x^(-0.31)~x, add=TRUE, col='red')
```
How about European metro?
```{r}
x = seq(1,14)
pop=c(11.02, 6.13, 6.03, 4.91, 4.59, 3.97, 3.57, 3.33, 3.19, 2.77, 2.66, 2.40, 2.09, 1.99)
plotPoints(log(pop) ~ log(x))
fitModel(log(pop)~A*log(x)+B)
plotPoints(pop~x)
exp(2.405931)
exp(2.41)
exp(2.406)
plotFun(exp(2.41)*x^(-0.61)~x, add=TRUE, col='red')
```



How about African metro?
```{r}
x = seq(1,14)
pop=c(20.90, 16.73, 14.37, 8.33, 7.77, 6.70, 6.58, 5.83, 5.78, 5.28, 5.20, 4.80, 4.62, 4.00)
plotPoints(log(pop) ~ log(x))
fitModel(log(pop)~A*log(x)+B)
exp(3.1471)
exp(3.15)
plotPoints(pop~x)
plotFun(exp(3.15)*x^(-0.65)~x, add=TRUE, col='red')
```


It has been long observed that the populations of the largest cities in a geographic region seem to follow a power law
$pop(x) = C x^p$ where $x$ is the rank of the city's population (1st, 2nd, 3rd, 4th largest, and so on). 

Here are the populations of the 14 largest metro areas in Africa, Europe and the United States.

africaPop = c(20.90, 16.73, 14.37, 8.33, 7.77, 6.70, 6.58, 5.83, 5.78, 5.28, 5.20, 4.80, 4.62, 4.00)
europePop = c(11.02, 6.13, 6.03, 4.91, 4.59, 3.97, 3.57, 3.33, 3.19, 2.77, 2.66, 2.40, 2.09, 1.99)
usPop = c(18.82, 12.46, 8.88, 6.49, 6.40, 6.17, 5.91, 5.73, 5.38, 4.58, 4.32, 3.53, 3.46, 3.31)

(a) Find the values of C and p for each of these three geographic regions. Enter them below

(b) Let's call two values a and b "similar" if their ratio is close to one (0.9 < a/b < 1.1). Otherwise they are dissimilar. How would you describe the set of constants that you found?

* The p's are similar and the C's are similar
* The p's are similar and the C's are dissimilar
* The p's are dissimilar and the C's are similar
* The p's are dissimilar and the C's are dissimilar

(c) Do your observations support the hypothesis that there is a "universal growth law" for metro areas? Why or why not?

(c) Do your values for $C$ and/or $p$ support the hypothesis that there is one "universal growth law" 

There are two possible

There is a universal growth rate for metro areas that applies to all geographic regions
Different geographic regions have very different growth rates. 

Compare the three values for $$p$$ that you found. 

Answer the following questions:
* What is the meaning of the power $$p$$ in this situation?
* Do your three values of $$p$$ results suggest that there is a "universal law" for the relative sizes of metro areas,  or that "local factors" play an important role? Explain.



##### Dimensional Analysis


A scientist wants to determine a formula for the resistive force $f$ experienced when an object moves through a fluid. Experiments suggest that this force depends upon
* the velocity $v$ that the object is moving
* the area $a$ of the cross-section of the object
* The density $\mu$ of the liquid.

This scientist has asked you to suggest a formula for $f(v,a,\mu)$. 

(This scientist will then test your proposed formula against their experimental data for various objects and fluids.)

###### Periodic

Prior to the introduction of a vaccine in 1968, measles outbreaks in the United States had a periodic behavior. The graph below shows the weakly cases of measles reported between 1948 and 1968. 

Let $M(t)$ be the number of weakly measels cases where $t$ is the number of **years** since January 1, 1950. Use this graph to model $M(t)$ using either a sine function or a cosine function. 



###### Rate of Change

Here is a link to the graph of the unemployment rate (%) in Minneapolis-St Paul Metro Area for January 1990 - August 2021.

https://fred.stlouisfed.org/series/MINN427URN

When you mouse over a particular point, it will give you the unemployment rate for the corresponding month.

1. Calculate the average rate of change of the unemployment rate between January 1990 and August 2021. Measure time in years (not months).
1. Calculate the average rate of change of the unemployment rate between November 2008 and November 2019. Once again, measure time in years (not months).
1. Does this graph show any year-to-year seasonal patterns to unemployment? Explain.



##### Linear Function

An apple farmer will grow crops whenever they can sell those crops for a profit at the local market. They are considering buying an orchard located $d$ miles from the market.
The market price for a bushel of apples is \$30. Meanwhile, it costs the farmer \$17 to produce those apples, and \$0.50 per mile to transport those apples to the market.

Find the distance maximum distance $M$ that the farmer would agree to buy a farm. (In other words, if $d < M$ then the farmer would buy the orchard.)

```{r}
(30-17)/.5=26
```




##### Drug Surge

Consider the concentration C(x,t) in micrograms per liter (mg/L), of a drug in the blood as a function of x, the amount, in mg, of the drug given and t, the time in hours since the injection. For 0 \leq x \leq 4 and t  \geq 0  , we have

plotFun(22*t*exp(-(5-x)*t),x&t, x.lim=c(0,30),t.lim=c(0,30))

```{r}
h = makeFun(22*t*exp(-(5-x)*t)~x&t)
h(0,0)
plotFun(1000* t*exp(-(5-x/1000)*t)~x&t, x.lim=c(0,4000),t.lim=c(0,4), levels=seq(0,1500,50), filled=FALSE, xlab='dosage (micrograms)', ylab='time (hours)',title='Concentration (micrograms per liter)')
```

On this domain, for a fixed TIME this function is  INCREASING/DECREASING/ INCREASING THEN DECREASING? DECREASING THEN INCREASING

On this domain, for a fixed DOSAGE this function is  INCREASING/DECREASING/ INCREASING THEN DECREASING? DECREASING THEN INCREASING

The drug is effective at concentrations of 150 micrograms/liter. What is the minimum dose that we should give if we want the drug to be at an effective level at least 1 hour? 2500, 3000, 3500, 4000





#### Exam 2


```{r}
h = makeFun(22*t*exp(-(5-x)*t)~x&t)
h(0,0)
plotFun(1000* t*exp(-(5-x/1000)*t)~x&t, x.lim=c(0,4000),t.lim=c(0,4), levels=seq(0,1500,50), filled=FALSE, xlab='dosage (micrograms)', ylab='time (hours)',title='Concentration (micrograms per liter)')
```


