# Partial Derivatives

```{r, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(mosaic))
suppressPackageStartupMessages(library(mosaicCalc))
```

## Activities

### Estimating From Data

The maximum duration of a scuba dive $T(V,D)$ (in minutes) depends on the volume $V$ of air (at sea-level pressure) in the tank and the depth $D$ (in feet) of the dive.
The following table shows the estimated dive times for various combinations of air volume and dive depth.

<table style="width: 70%; margin-left: auto; margin-right: auto;" border="1" cellspacing="2" cellpadding="2" align="center">
<tbody>
<tr>
<td style="margin-left: 30px; text-align: center;" rowspan="2" colspan="2"><strong>Dive Time (minutes)</strong> </td>
<td style="text-align: center;" colspan="4"><strong> Air Volume</strong></td>
</tr>
<tr>
<td style="text-align: center;"><strong>20 ft<sup>3</sup></strong></td>
<td style="text-align: center;"><strong>40 ft<sup>3</sup></strong></td>
<td style="text-align: center;"><strong>60 ft<sup>3</sup></strong></td>
<td style="text-align: center;"><strong>80 ft<sup>3</sup></strong></td>
</tr>
<tr align="right">
<td style="margin-left: 30px; text-align: center;" rowspan="5"><strong>Depth</strong></td>
<td style="text-align: center;"><strong>20 ft</strong></td>
<td style="text-align: center;">12.5</td>
<td style="text-align: center;">25.0</td>
<td style="text-align: center;">37.5</td>
<td style="text-align: center;">50.0</td>
</tr>
<tr align="right">
<td style="text-align: center;"><strong>40 ft</strong></td>
<td style="text-align: center;">9.0</td>
<td style="text-align: center;">18.0</td>
<td style="text-align: center;">27.0</td>
<td style="text-align: center;">36.0</td>
</tr>
<tr align="right">
<td style="text-align: center;"><strong>60 ft</strong></td>
<td style="text-align: center;">7.1</td>
<td style="text-align: center;">14.2</td>
<td style="text-align: center;">21.3</td>
<td style="text-align: center;">28.4</td>
</tr>
<tr align="right">
<td style="text-align: center;"><strong>80 ft</strong></td>
<td style="text-align: center;">5.8</td>
<td style="text-align: center;">11.6</td>
<td style="text-align: center;">17.4</td>
<td style="text-align: center;">23.2</td>
</tr>
</tbody>
</table>

Use this table to estimate the derivatives
$\displaystyle{\frac{\partial T}{\partial V}}$ and $\displaystyle{\frac{\partial T}{\partial D}}$ at the point $(V=60, T=40)$. Give units for each of these derivatives and then interpret what these numbers mean.

### Estimating From a Contour Plot

The production function $F(L,K)$ for a factory gives the number of units produced by labor $L$ and capital $K$ (both measured in dollars). This production function is given by
$$
F(L,K) = 10 L^{2/7}K^{5/7}.
$$
Here is a contour plot of the production function.


```{r, echo=FALSE, fig.asp=1, include=FALSE}


plotFun(10* x^{2/7} * y^{5/7}~x&y, xlim=c(0,1000), ylim=c(0,1000), levels = seq(1000,9000,1000), filled = FALSE, xlab="labor L", ylab = "capital K", lwd = 3)
ladd( panel.abline(h=seq(100,900,100),v=seq(100,900,100),col='gray50') )
```

```{r, echo=FALSE, fig.asp=1}

ladd( panel.abline(h=seq(100,900,100),v=seq(100,900,100),col='gray50') )
```

1. Estimate the production output for $L=400$ and $K=600$.
2. Estimate the partial derivative $\displaystyle{\frac{\partial F}{\partial L}}$ at $(400,600)$.
3. Estimate the partial derivative $\displaystyle{\frac{\partial F}{\partial K}}$ at $(400,600)$.



### Estimating Using Desmos

Use desmos to define the prodution function $F(L,K) = L^{2/7}  K^{5/7}$ from the previous problem and to answer the analogous questions

1. Calculate the production outpuut $F(400,600)$.
2. Estimate the partial derivative $\displaystyle{\frac{\partial F}{\partial L}}$ at $(400,600)$. Your answer must be correct to two decimal places.
3. Estimate the partial derivative $\displaystyle{\frac{\partial F}{\partial K}}$ at $(400,600)$.
Your answer must be correct to two decimal places.

Compare your three answers to the estimates from the contour plot in the previous problem.


## Solutions

### Estimating From Data

We estimate the partial derivatives. We have
$$
T_V(60,40) \approx \frac{T(80,40) - T(60,40)}{80-60}
= \frac{36 - 27}{80-60} = \frac{9}{20} = 0.45 \mbox{ min/ft}^3
$$
where we have used the interval $60 \leq V \leq 80$ for our estimate. We could also use $40 \leq V \leq 80$ or $40 \leq V \leq 80$ as the interval for our estimate.
$$
T_D(60,40) \approx \frac{T(60,60) - T(60,40)}{60-40}
= \frac{21.3 - 27}{60-40} = - \frac{5.7}{20} = -0.285 \mbox{ min/ft}
$$
where we have used the interval $40 \leq D \leq 60$ for our estimate. We could also use $20 \leq D \leq 40$ or $20 \leq D \leq 60$ as the interval for our estimate.


The units for $T_V(V,D)$ are minutes per cubic feet. An  additional cubic foot of air will allow you to dive for an extra 0.45 minutes.

The units for $T_V(V,D)$ are minutes per feet. An  additional foot of dive depth will reduce your dive time: the change is $-0.285$ minutes.

### Estimating From a Contour Plot

We have
$$
F(400,600) \approx 5300.
$$

Let's estimate by using the contour line $F(L,K) = 5000$. I estimate that $F(325,600)=5000$ and that $F(400,550)=5000$. Therefore

$$
F_L(400,600) \approx \frac{5300 - 5000}{400 - 325} = \frac{300}{75} = 4
$$
and
$$
F_K(400,600) \approx \frac{5300 - 5000}{600-550}
= \frac{300}{50} = 6.
$$

### Estimating Using Desmos

Here is a [link to my desmos calculations](https://www.desmos.com/calculator/piexczmrsn).

1. $F(400,600) = 534.37$
1. $F_L(400,600) = 0.38$
1. $F(400,600) = 0.64$
