# The Gradient


## Activities


### Partial Derivative Practice

Find the partial derivatives $f_x$ and $f_y$ for each of these functions.

1.  $f(x,y) = x^5 + x^2y^3 + y^4$
1. $f(x,y) = e^x \sin(y) + \ln(y) \cos(x)$
1. $f(x,y) = \ln (x^2 - y^2)$
1. $f(x,y) = x^2 y^3 e^x \sin(y)$

### Second Partial Derivative Practice

For each of these functions, find the second partial derivatives $f_{xx}, f_{yy}, f_{xy}$ and $f_{yx}$. Confirm that $f_{xy} = f_{yx}$

1. $f(x,y) = 7x^3 + \cos(4y)$
2. $f(x,y) = e^{2x+3y}$

### Gradient Practice

Calculate the gradient $\nabla f(x,y)$ for each of these functions

1. $f(x,y) = 8x^2y^4$
1. $\displaystyle{f(x,y) = \frac{x}{y}}$
1. $f(x,y) = \ln(xy)$

### Gradient on a Contour Diagram

Here is a contour plot of $f(x,y)$. At each of the three points indicated by $*$, draw a vector in the direction of the gradient $\nabla f$. 

```{r, echo=FALSE, fig.keep = "last"}
P = makeFun(2*(x-4)^2 - 2*(x-4)*(y-3) - 3*(y-3)^2 ~ x&y)


plotFun(P(x,y) ~ x&y, filled=FALSE, x.lim=range(0,7),y.lim=range(0,7), levels =seq(-50,50,10) )
plotPoints(c(5,2,6)~c(0.9,2.2,4.5), add=TRUE, col='blue', pch=8)
```

### Gradient and the Direction of Greatest Increase

Let $f(x,y) = 3x^2 - 4xy + y^2$.

1. Find the gradient vector $\nabla f(2,1)$ the at the point $(2,1)$. 
1. Find the rate of change of $f(x,y)$ at $(2,1)$ in the direction of greatest increase.



## Solutions


### Partial Derivative Practice


1.  $f(x,y) = x^5 + x^2y^3 + y^4$

$$
f_x = 5x^2 + 2xy^3 \qquad \qquad f_y = 3x^2y^2 + 4y^3
$$

1. $f(x,y) = e^x \sin(y) + \ln(y) \cos(x)$

$$
f_x = e^x \sin(y) - \ln(y) \sin(x) \qquad \qquad f_y = e^x \cos(y) + \frac{1}{y} \cos(x)
$$

2. $f(x,y) = \ln (x^2 - y^2)$


$$
f_x = \frac{2x}{x^2-y^2} \qquad \qquad f_y = - \frac{2y}{x^2-y^2}
$$

3. $f(x,y) = x^2 y^3 e^x \sin(y)$

Let's rewrite this function as $x^2 e^x y^3 \sin(y)$

$$
f_x = (2xe^x + x^2e^x) y^3\sin(y) \qquad \qquad
f_y = (3y^2 \sin(y) - y^3 \cos(y)) x^2 e^x
$$

### Second Partial Derivative Practice


1. $f(x,y) = 7x^3 + \cos(4y)$

\begin{align*}
f_x &= 21 x^2 \\
f_{xx} &= 42 x \\
f_{xy} &= 0 \\
\\
f_y &= -4 \sin(4y) \\
f_{yy} &= -16 \cos(y) \\
f_{yx} &=0
\end{align*}


2. $f(x,y) = e^{2x+3y}$

\begin{align*}
f_x &= 2e^{2x+3y} \\
f_{xx} &= 4e^{2x+3y} \\
f_{xy} &= 6e^{2x+3y} \\
\\
f_y &= 3e^{2x+3y} \\
f_{yy} &= 9e^{2x+3y} \\
f_{yx} &= 6e^{2x+3y}
\end{align*}

### Gradient Practice

Calculate the gradient $\nabla f(x,y)$ for each of these functions

1. $\nabla f(x,y) = \langle 16x y^4, 32x^2y^3 \rangle$
1. $\nabla f(x,y) = \langle y^{-1}, -xy^{-2} \rangle$ 
1. $f(x,y) = \langle x^{-1}, y^{-1} \rangle$

### Gradient on a Contour Diagram

Here is a contour plot of $f(x,y)$. At each of the three points indicated by $*$, draw a vector in the direction of the gradient $\nabla f$. 

```{r, echo=FALSE, fig.keep = "last"}
P = makeFun(2*(x-4)^2 - 2*(x-4)*(y-3) - 3*(y-3)^2 ~ x&y)


plotFun(P(x,y) ~ x&y, filled=FALSE, x.lim=range(0,7),y.lim=range(0,7), levels =seq(-50,50,10) )
plotPoints(c(5,2,6)~c(0.9,2.2,4.5), add=TRUE, col='blue', pch=8)

```

### Gradient and the Direction of Greatest Increase

1. $\nabla f(x,y) = \langle 6x - 4y, -4x + 2y \rangle$ so 
$\nabla f(x,y) = \langle 12 - 4, -8 + 2 \rangle = \langle 8, -6 \rangle$.
1. $\| \nabla f(2,1) \| = \sqrt{8^2 + (-6)^2} = \sqrt{64+36} = \sqrt{100} = 10.$

